---
title: "Demo"
author: "Seraphina (Junming) Shi"
output:
  html_document:
    toc: true
    toc_float: true
---

Based on the *hal9001* R package by Rachael Phillips, Jeremy Coyle, and Nima Hejazi.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives
By the end of this demo, you will be able to:  

* Fit HAL with CV-selected penalty or undermoothed HAL.  
* Interpret the results from HAL-fit.   
* Estimate the target parameter with HAL as a working model and obtain delta method based confidence intervals.  
  + Average treatment effects.  
  + Conditional average treatment effects.  
  + Dose response curves.  

# 1. Introduction 

# 2. How to Fit HAL
In this section, the functions of fitting HAL is demonstrated. 

**Fitting any HAL consists of the following 2 steps**:  
1. Define the outcome vector, Y, and the covariates matrix X.  
2. Fit the HAL with selected smoothness order and other parameters.    

**Running example with WASH Benefits dataset**:  
We will use the WASH Benefits Bangladesh study as an example to guide this demo of HAL-based plug-in estimator. 

#### Preliminaries
First, we need to load the data, relevant functions, and relevant packages into the R session, and then make sure there is no missing values in the dataset. 


**Load the data**
```{r}
library("data.table")

washb_data <- fread(
  paste0(
    "https://raw.githubusercontent.com/tlverse/tlverse-data/master/",
    "wash-benefits/washb_data.csv"
  ),
  stringsAsFactors = TRUE
)
```

Take a peek at the first few rows of WASH dataset:
```{r}
head(washb_data)
```

**Remove missing values**
Here we simply keep the complete cases by removing the rows with missing values. You can choose any other methods to deal with missing values.
```{r}
washb_data <- washb_data[complete.cases(washb_data), ]
```

#### Install hal9001 pacakge (as needed)
To install any packge in R, we recommend clearing the R workspace and subsequently restarting the R session. In RStudio, you can accomplish this by navigating to the "Session" tab, selecting "Clear Workspace," and then proceeding to click on "Session" again and choosing "Restart R."

```{r, eval = FALSE}
install.packages("hal9001")
```

#### Load hal9001 package 
Once the *hal9001* package is installed, we can load itas other other R package:
```{r}
library(hal9001)
```

load the *'dplyr'* package for easy and quick data manipulation.    
load the *'glmnet'* package used in the undersmoothing HAL process. 
```{r}
library(dplyr)
library(glmnet)
```

#### Source relevant functions
You need to download the *'functions_UHAL_deltaCI.R'* for the github repo. Then source it to load relevant functions for fitting Undermoothed HAL and obtainning confidence intervals using the delta method.
```{r}
source('functions_UHAL_deltaCI.R')
```

## 2.1 Prepare data

First, identify the outcome variable and save it as a numeric vector.  
```{r}
Y <- as.numeric(washb_data$whz)
```

Then, identify the treatment variable and other covariates, save them as a numeric matrix. If the treatment or other variables are categorical or ordinal, we need to convert them into numeric or dummy variables.   

```{r}
x_names = c("tr", 
            "fracode", "month", "aged", "sex", "momage", "momedu", "momheight", 
            "hfiacat", "Nlt18", "Ncomp", "watmin", "elec", "floor", "walls", 
            "roof", "asset_table", "asset_chair", "asset_khat", 
            "asset_chouki", "asset_tv", "asset_refrig")

X <- washb_data %>% 
    select(all_of(x_names)) %>% 
    mutate_if(sapply(., is.factor), as.numeric)
```

We have categorical treatment variable initially. To be prepared to fit HAL, we transformed it into numeric values. Here is the transformation mapping: 
```{r}
table(X$tr, washb_data$tr)
```

## 2.2 Fit HAL
### 2.2.1 CV-HAL

To fit the CV-HAL using the prepared X and Y above, we call the '*fit_hal*' function from the *hal9001* package. The following code fits HAL with the default setting, where max_degree = 1, smoothness_orders = 1, and other arguements. You can look for '*fit_hal*' in the help pane in Rstudio or go to *https://rdrr.io/cran/hal9001/man/fit_hal.html* for more information. Before fitting, we will set a random number generator so the results are reproducible, and we will also time it. 
```{r}
start_time <- proc.time() # start time

set.seed(123)
# default setting 
hal_CV <- fit_hal(X = X, Y = Y, 
                  family = "gaussian")

runtime_hal_fit <- proc.time() - start_time # end time - start time = run time
runtime_hal_fit
```

Now we try to fit HAL with '*use_min = FALSE*' and zero smoothness order, where "lambda.1se" is used and the basis functions will all be indicators. To enable the downstream confidence interval calculation, we need to set the arguement '*return_x_basis = T*'. 
```{r}
start_time <- proc.time() # start time

set.seed(123)
hal_fit_zeroorder <- fit_hal(X = X, Y = Y, 
                             family = "gaussian", 
                             fit_control = list(use_min = FALSE),
                             smoothness_orders = 0,
                             return_x_basis = T)

runtime_hal_fit <- proc.time() - start_time # end time - start time = run time
runtime_hal_fit
```

### 2.2.2 Undersmoothed HAL
First, we need to fit a cv-hal with the arguement '*return_x_basis = T*' by calling the '*fit_hal*' function. Then we pass in *'X'*, *'Y'*, *'hal_cv' fit object*, and *'family'* to the '*undersmooth_hal*' function (from the *'functions_UHAL_deltaCI.R'* file) to get the undersmoothed lambda. Then fit the undersmoothed HAL by calling *fit_hal*' function again with the undersmoothed lambda and other arguements same as initially fitting the cv-hal. 

```{r}
start_time <- proc.time() # start time

set.seed(123)
hal_cv <- fit_hal(X = X, Y = Y, 
                  smoothness_orders = 1,
                  family = "gaussian",
                  return_x_basis = TRUE)

undersmooth_hal_results <- undersmooth_hal(X, Y, 
                                           fit_init = hal_cv,
                                           family = "gaussian")

hal_u <- fit_hal(X = X, Y = Y, 
                smoothness_orders = 1,
                family = "gaussian",
                return_x_basis = TRUE,
                fit_control = list(cv_select = FALSE),
                lambda = undersmooth_hal_results$lambda_under)

runtime_uhal_fit <- proc.time() - start_time # end time - start time = run time
runtime_uhal_fit
```

There can be situations that there is no need for undersmoothing or we cannot undersmooth further. We can check this by whether '*lambda_init*' and '*lambda_under*' in the results of calling '*undersmooth_hal*' function are the same. If this is the case, we can just proceed with '*hal_cv*'. 

```{r}
set.seed(123)
hal_cv_secondorder <- fit_hal(X = X, Y = Y, 
                    smoothness_orders = 2,
                    family = "gaussian",
                    return_x_basis = TRUE)

undersmooth_hal_results <- undersmooth_hal(X, Y, 
                                           fit_init = hal_cv_secondorder,
                                           family = "gaussian")

print(undersmooth_hal_results$lambda_init == undersmooth_hal_results$lambda_under)
```

## 2.3 Sumarizing HAL fits

The summary of a hal fit object will have the table of coefficients and the corresponding basis functions with non-zero coefficients. The lambda used will also be stored here.

```{r}
hal_fit_summary <- summary(hal_u)
print(hal_fit_summary)

print(hal_fit_summary$lambda)
```



# 3. Obtainning Estimates 
## 3.1 Predictions
### 3.1.1 Predicitons with new data
### 3.1.2 Conterfactual predictions
## 3.2 ATE
## 3.3 CATE
## 3.4 Dore response curve

# 4. Confidence Intervals using Delta Method
## 4.1 Predictions
### 4.1.1 Binary outcome
### 4.1.2 Continuous outcome
## 4.2 ATE
## 4.3 CATE
## 4.4 Dore response curve

# 5. Dore Response Curve Example with a Simulated Dataset
## 5.1 Fit HAL
## 5.2 Dose response curve estimatation
## 5.3 Confidence intervals

